Lossy compression of DNA sequencing quality data

### 2015-05-01 manuscript/q-compression.tex
some small edits

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each
class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated
the effects of lossy compression on identifying variants within a data
set. We build on these prior works and access the effects of
% quantifying compression efficiency, we also assess the effects of
quality value information loss resulting from compression on
additional subsequent genomic analyses including read preprocessing
(filtering and trimming), genome assembly, and read mapping.

@@

\textsc{QualComp} is a lossy compression tool that models quality
values as a multivariate Gaussian distribution, computing the mean and
covariance for each position in the read\citep{Ochoa:2013rt}. Once the
model parameters are calculated, they are stored by the decoder to
later reconstruct a \emph{representative} quality value.
\textsc{QualComp} takes as input a user-specified rate (bits per read)
and then poses an optimization problem of how to allot these bits for
a given position within the read while minimizing the \textsc{mse}. The quality values
can be clustered before hand to produce more accurate models.

@@

\begin{figure*}[!tb]%figure2
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponding to original
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

@@

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the sequence and uses
a sliding window ($0.1 \times$ the read length) to find locations
where the window's average quality falls below a given threshold (20
by default). The sequence is then trimmed from this position to the
respective end.
%uses a variable sliding window approach where base qualities are
%averaged over a window ($0.1 \times$ the read length) and trims from
%the position where the window's average quality falls below a given
%threshold (20 by default).
If the trimmed sequence length is less than a certain threshold (20 by
default), then the sequence is discarded.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
outperforms all others in all metrics (Table
\ref{fig:assembly_ranks}).
%The assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set
%using the uncompressed reads outperforms all compression methods in
%terms of \textsc{lap}, NG50, relocations, translations, and
%inversions.
Among the compression methods, the profile encoding performed best,
polynomial regression modeling with degree $<=$ 7 performed worst, and
other methods were intermediate (Fig.~\ref{fig:assembly_ranks}).

@@

\begin{figure}[!tbp]
%\centerline{\includegraphics[width=3.65in]{mse_frag.png}}
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}

>>>>

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective some loss of information is
deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop a robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and consider an exemplar of each
class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated
the effects of lossy compression on identifying variants within a data
set. We build on these prior works, and access the effects of quality
value information loss resulting from compression on subsequent
genomic analyses including read preprocessing (filtering and
trimming), genome assembly, and read mapping.

@@

@@

\begin{figure*}[!tb]%figure2
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponding to original
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

@@

\subsection{Comparison to other methods}

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool that models quality values as a multivariate
Gaussian distribution, computing the mean and covariance for each
position in the read\citep{Ochoa:2013rt}. Once the model parameters
are calculated, they are stored by the decoder to later reconstruct a
\emph{representative} quality value.  \textsc{QualComp} takes as input
a user-specified rate (bits/read) and then poses an optimization
problem of how to allot these bits for a given position within the
read while minimizing the \textsc{mse}. The quality values can be
clustered before hand to produce more accurate models.

@@

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the sequence and uses
a sliding window ($0.1 \times$ the read length) to find locations
where the window's average quality falls below a given threshold (20
by default). The sequence is then trimmed from this position to the
respective end. If the trimmed sequence length is less than a certain
threshold (20 by default), then the sequence is discarded.

@@

No assembly of the \textit{Rhodobacter sphaeroides} 2.4.1 data set was
best in all metrics (Table \ref{fig:assembly_ranks}). Among the
compression methods, the profile encoding performed best, polynomial
regression modeling performed worst, and other methods were
intermediate (Fig.~\ref{fig:assembly_ranks}).

@@

\begin{figure}[!tbp]
\centerline{\includegraphics[width=3.65in]{rhodo_assembly_results.pdf}}
\caption{Rankings of compression methods based on \textit{Rhodobacter
    sphaeroides} assembly attributes sorted by overall
  rank. Assemblies were constructed using \textsc{allpaths-lg}.
  Rankings above the median value are in cyan, those below the median
  value in magenta.}
  \label{fig:assembly_ranks}
\end{figure}
